---
title: "Project1"
format: html
editor: visual
---

### Data

The files to be read contain information from the *Census Bureau* in csv format. The aim is to create generic functions that can process and plot files in specific tabular format. Firstly, data processing steps will be applied to [File 1](https://www4.stat.ncsu.edu/~online/datasets/EDU01a.csv).

### Data Processing

#### First Steps

The first step requires retuning specific columns from full set. This is made easier using the `tidyverse` packages. Loading the `tidyverse` packages using `require()` .

```{r}
require(tidyverse)
```

Reading in [File 1](https://www4.stat.ncsu.edu/~online/datasets/EDU01a.csv) using `read_csv().`

```{r}
file_1 <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/EDU01a.csv")

head(file_1) #Preview
```

Selecting the required columns with `select()` from `dplyr`. Selecting the columns `Area_name`, `STCOU` and columns ending with "D" (using `ends_with()`). Renaming `Area_name` to `area_name` with `mutate()` .

```{r}
file_1 |>
  mutate(area_name = Area_name) |>
  select(area_name,STCOU,ends_with("D"))
```

Template Leftover\_\_\_\_\_\_\_\_\_\_\_\_\_\_

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
